A Lanczos Procedure for Approximating Eigenvalues of Large Stochastic Matrices

### 2013-10-18 tex/InvariantSubspaces.tex
initial commit of tex source for nyu thesis style setup

%\documentclass[11pt]{amsart}
\documentclass{au}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% For completion by editorial office:
 \presentedby{\dots}
 \received{\dots}{\dots}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage[colorlinks=true,urlcolor=black,linkcolor=black,citecolor=black]{hyperref}

\usepackage{amsmath,amssymb,url,mathrsfs}
\usepackage{enumerate}
\usepackage{tikz}
\usetikzlibrary{calc}
\usepackage{scalefnt}
\numberwithin{equation}{section}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{notation}[theorem]{Notation}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}[theorem]{Example}
\newtheorem*{remarks}{Remarks}

\newcommand{\todo}[1]{~\\[5pt]{\bf TODO(wjd):} #1\\[5pt]}
%\newcommand{\suchthat}{\ensuremath{:}}    % (old notation)
\newcommand{\suchthat}{\ensuremath{\mid}}  % (new notation)
%\newcommand{\defeq}{\ensuremath{:=}}      % (old notation)
\newcommand{\defeq}{\ensuremath{=}}        % (new notation)
\newcommand{\<}{\ensuremath{\langle}}
\renewcommand{\>}{\ensuremath{\rangle}}
\newcommand{\beps}{\ensuremath{\boldsymbol{\varepsilon}}}
\newcommand{\bN}{\ensuremath{\mathbf{N}}}
\newcommand{\bA}{\ensuremath{\mathbf{A}}}
\newcommand{\bB}{\ensuremath{\mathbf{B}}}
\newcommand{\bBi}{\ensuremath{\mathbf{B}_i}}
\newcommand{\sB}{\ensuremath{\mathcal{B}}}
\newcommand{\sC}{\ensuremath{\mathcal{C}}}
\newcommand{\sS}{\ensuremath{\mathscr{S}}}
\newcommand{\sT}{\ensuremath{\mathscr{T}}}
\newcommand{\sI}{\ensuremath{\mathcal{I}}}
\newcommand{\sE}{\ensuremath{\mathcal{E}}}
\newcommand{\sO}{\ensuremath{\mathcal{O}}}
%\newcommand{\id}{\ensuremath{\mathrm{id}}}
\DeclareMathOperator{\id}{id}
%\newcommand{\Eq}{\ensuremath{\mathrm{Eq}}}
\DeclareMathOperator{\Eq}{Eq}
%\newcommand{\Cg}{\ensuremath{\mathrm{Cg}}}
\DeclareMathOperator{\Cg}{Cg}
%\newcommand{\Con}{\ensuremath{\mathrm{Con\,}}}
\DeclareMathOperator{\Con}{Con}
%\newcommand{\Sub}{\ensuremath{\mathrm{Sub}}}
\DeclareMathOperator{\Sub}{Sub}
%\newcommand{\Pol}{\ensuremath{\mathrm{Pol}}}
\DeclareMathOperator{\Pol}{Pol}
%\newcommand{\Clo}{\ensuremath{\mathrm{Clo}}}
\DeclareMathOperator{\Clo}{Clo}
\newcommand{\ps}[1]{\ensuremath{^{(#1)}}}
\newcommand{\piB}{\ensuremath{\pi_B}}
\newcommand{\hpsi}{\ensuremath{\widehat{\psi}}}
\newcommand{\htheta}{\ensuremath{\widehat{\theta}}}
\newcommand{\supi}{\ensuremath{^{i}}}
\newcommand{\supj}{\ensuremath{^{j}}}
\renewcommand{\leq}{\ensuremath{\leqslant}}
\renewcommand{\nleq}{\ensuremath{\nleqslant}}
\renewcommand{\geq}{\ensuremath{\geqslant}}
\newcommand{\meet}{\ensuremath{\wedge}}
\newcommand{\join}{\ensuremath{\vee}}
\newcommand{\Meet}{\ensuremath{\bigwedge}}
\renewcommand{\Join}{\ensuremath{\bigvee}}
\newcommand{\resB}{\ensuremath{|_{_B}}}
\newcommand{\resBi}{\ensuremath{|_{_{B_i}}}}
\newcommand{\eps}{\ensuremath{\varepsilon}}
\newcommand{\hatmap}{\ensuremath{\widehat{\phantom{x}}}}
\newcommand{\one}{\ensuremath{\mathbf{1}}}
\newcommand{\two}{\ensuremath{\mathbf{2}}}
\newcommand{\three}{\ensuremath{\mathbf{3}}}
\newcommand{\tbeta}{\ensuremath{\widetilde{\beta}}}
\newcommand{\ttheta}{\ensuremath{\widetilde{\theta}}}
\newcommand{\hbeta}{\ensuremath{\widehat{\beta}}}
\newcommand{\cick}{\ensuremath{\sC_i}}
%\newcommand{\CE}{\ensuremath{\sC_r^{\sE}}}
\newcommand{\CE}{\ensuremath{\sB_r^n}}
\newcommand{\CO}{\ensuremath{\sC_r^{\sO}}}
\newcommand{\CICK}{\ensuremath{c_i/\beta \cup c^{K+1}_i/\beta^{\bB_{K+1}}}}
\newcommand{\GAP}{{\small GAP}}
\newcommand{\dotsize}{.8pt}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%  FRONT MATTER                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title[Approximating Eigenvalues of Large Stochastic Matrices]{A Lanczos Procedure for Approximating Eigenvalues of Large Stochastic Matrices}

%% First author (Note: The order of the items here is important!)
\author{William DeMeo}
\email{williamdemeo@gmail.com}
\urladdr{http://www.math.sc.edu/~demeow}
\address{Department of Mathematics\\
University of South Carolina\\Columbia 29208\\USA}

\thanks{The author wishes to thank a number of people who contributed to this
  work.  In particular, the primary referee made many insightful recommendations
  which led vast improvements, not only in the overall exposition, but also in
  the specific algebraic constructions, theorem statements, and proofs.
  This article is dedicated to Ralph Freese and Bill Lampe, who declined to be
  named as co-authors despite their substantial contributions to the project.}

%% Dedication (Optional)
\dedicatory{This article is dedicated to Ralph Freese and Bill
  Lampe.}

%% AMS subject classification; see http://www.ams.org/msc
%% Only one Primary. Possibly several Secondary.
\subjclass[2010]{Primary: 08A30; Secondary: 08A60, 06B10.} %, 06B15.}

%% Keywords and phrases
\keywords{congruence lattice, finite algebra, finite lattice
  representations}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%  MAIN MATTER                                                  %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstract}
The rate at which a Markov chain converges to a given probability distribution
has long been an active area of research. Well known bounds on this rate of
convergence involve the subdominant eigenvalue of the chain’s underlying
transition probability matrix. However, many transition probability matrices are
so large that we are unable to store even a vector of the matrix in fast
computer memory. Thus, traditional methods for approximating eigenvalues are
rendered useless.

In this paper we demonstrate that, if the Markov chain is reversible, and we
understand the structure of the chain, we can derive the coefficients of the
traditional Lanczos algorithm without storing a single vector. We do this by
considering the variational properties of observables on the chain’s state
space. In the process we present the classical theory which relates the
information contained in the Lanczos coefficients to the eigenvalues of the
Markov chain. 
\end{abstract}

\maketitle

%%%%%%%%%%%%%%%%%%%%%
\chapter{Introduction}
The rate at which a Markov chain converges to a given probability distribution has long
been an active area of research. This is not surprising considering this problem’s relevance to the ar-
eas of statistics, statistical mechanics, and computer science. Markov Chain Monte Carlo (MCMC)
algorithms provide important examples. These algorithms come in handy when we encounter a
complicated probability distribution from which we want to draw random samples. In statistical
mechanics, we might wish to estimate the phase average of a function on the state space. Goodman
and Sokal [6] examine Monte Carlo methods in this context. Examples from statistics occur in the
Bayesian paradigm when we are forced to simulate an unwieldy posterior distribution (see, e.g.,
Geman and Geman

To implement the MCMC algorithm, we invent a Markov chain that converges to the
desired distribution (this is often accomplished using the Metropolis algorithm
described in Chapter 5). Realizations of the chain will eventually represent
samples from this distribution. Sometimes ``eventually'' -- meaning all but
finitely many terms of the chain -- is just not enough. We need more practical
results. In particular, we want to know how many terms of the chain should be 
discarded before we are sampling from a distribution that is close (in total variation distance) to
% 03.txt
the distribution of interest. This is the purpose of bounding convergence rates for Markov chains.

Often the Markov chains encountered in this context satisfy a condition known in the
physics literature as detailed balance. Probabilists call chains with this property reversible. This
simply means that the chain has the same probability law whether time moves
forward or 
backward.\footnote{This is not a precise definition. In particular the chain must
  have started from its stationary distribution. Full rigor is postponed until Section 2.1.}
In this paper, we consider the rate at which such chains converge to a \emph{stationary distribution}.\footnote{This and other italicized terms are defined in Section 2.1.}

There are a number of different methods in common use for bounding convergence rates of
Markov chains, and a good review of these methods with many references can be found in  More
recently developed methods, employing logarithmic Sobolev inequalities, are reviewed in  Most
of the bounds in common use involve the sub—dominant eigenvalue of the Markov chain's transition
probability matrix, and thus require good approximations to such eigenvalues. In many applications,
however, the transition probability matrix is so large that it becomes impossible to store even a
single vector of the matrix in conventional computer memory. These so called out-of-core problems
are not amenable to traditional eigenvalue algorithms\footnote{By ``traditional
  eigenvalue algorithms'' we refer to those found, for example, in Golub and Van
  Loan[5]. See also the book by Demmel [1] for a more recent discussion.}
without modification. This paper develops 
such a modification for the Markov chain eigenvalue problem. In particular it develops a method
for approximating the first few eigenvalues of a transition probability matrix when we know the
general structure of the underlying Markov chain. The method does not require storage of large
matrices or vectors. Instead we need only simulate the Markov chain, and conduct a statistical
analysis of the simulation.

Here is a look at what follows. Section 2.1 contains a review of the relevant Markov chain
theory. Readers conversant in the asymptotic theory of Markov chains might wish to at least skim
Section 2.1, if only to become familiar with our notation. Section 2.2 describes functions on the state
% 04.txt
space of the Markov process. This section and Chapter 3 develop the context in which we formulate
the new ideas of the paper. In the last section of Chapter 3, Section 3.3, we present the familiar
Krylov subspace and explain why this represents our best approximation to a subspace containing
extremal eigenvectors of the transition probability matrix.\footnote{or, more
  precisely, a similarity transformation of this matrix.} The first section of
Chapter 4 describes the \emph{Lanczos algorithm} for generating an orthonormal basis
for the Krylov subspace. As it stands, this algorithm is useless for an
out-of-core problem such as ours since, by definition of such problems, 
it requires too much data movement; all the computing time is spent swapping data between slow
and fast memory (e.g. between the hard disk and cache). Therefore, we discuss alternatives to
Lanczos and demonstrate that the \emph{Lanczos coeficients} are readily available through simulations of
the Markov chain, which fact allows us to avoid the standard algorithm altogether. Following this
is a chapter describing the Metropolis algorithm used to produce a reversible stochastic matrix.
It is here that we experiment with the procedure described in Section 4.2 and approximate the
extremal eigenvalues of the matrix, without storing any of its vectors. Finally, Chapter 6 concludes
the paper.
% 05.txt

\chapter{Markov Chains}

\section{General Theory}

This review of Markov chain theory can be found in any good probability text. The present
discussion is most similar to that of Durrett [3], to which we refer the reader desiring greater detail.

\subsection{The Basic Setup}

Heuristically, a Markov chain is a stochastic process with a lack of memory property. Here
this means that the future of the process, given its past behavior and its present state, depends only
on its present state. This is the probabilistic analogue of a familiar property of classical particle
systems. Given the position and velocities of all particles at time t, the equations of motion can be
completely solved for the future evolution of the system. Thus, information describing the behavior
of the process prior to time t is superﬂuous. To be a bit more precise, if technical, we need the
following definitions.

\bibliographystyle{spmpsci}
\bibliography{wjd}

\end{document}

%%%%%%%%%%%%%%%%%%%%%  END DOCUMENT %%%%%%%%%%%%%%%%%%%%%%%%%

Definition 2.1.1 Let (S, 8) be a measurable space. A sequence X,., n 2 0, of random variables

taking values in S is said to be a Markov chain with respect to the filtration a(Xo, . . . ,X,,) if for
% 06.txt
a.llBES,

p(x,,+1 e B | a(Xo, . . . ,X,.)) = P(X,.+1 e B 1 o(X,,)) (2.1)

Equation (2.1) merely states that, given the present location of X,,, the past is irrelevant for

predicting the location of X,,+1.

Definition 2.1.2 A function p : S x S —> R is said to be a transition probability if:
1. for each :1: E S, A —> p(a:, A) is a probability measure on (S, S).
2. for each A E 8, :1: —> p(z, A) is a measurable function.

We call X,, a Markov chain with transition probabilities 12,, if
P(Xn+l E B l a(Xn)) =Pn(XnaB) (22)

The spaces (S, S) that we encounter below are standard Borel spaces, so the existence of the
transition probabilities follows from the existence of regular conditional probabilities on Borel spaces
— a standard measure theory result (see e.g. [3] page 230).

Suppose we are given an initial probability distribution p on (S, S) and a sequence p,, of

transition probabilities. We can define a consistent set of finite dimensional distributions by

P<X,- e B,-,0 s 1 s n) = / #(d1‘o) / Po($o.d-'51) . .. / Pn-l(97n—1ad$n) (2.3)
B0 B1 B1:
Furthermore, denote our probability space by
(9,?) = (S{0,1,...},8{0,l,...})

We call this sequence space and it is defined more explicitly by

sf°»1=-~-} = {(wo,w1,...) :w.- e S}

S{°*1""} = ¢7(w :40; E A,- E 3)

% 07.txt
The Markov chain that we will study on this space is simply X" (w) = w,., the coordinate maps.
Then, by the Kolmogorov extension theorem, there exists a unique probability measure P,, on
(9,?) so that the X" (w) have finite dimensional distributions (2.3).

If instead of p, we begin with the initial distribution 6,, i.e. point mass at 2:, then we
denote the probability measure by P3. With such measures defined for each 2:, we can in turn

define distributions P,,, given any initial distribution p, by

PM = / u<da.->P.(A>

That the foregoing construction — which, recall, was derived merely from an initial distri-
bution p and a sequence p,, of transition probabilities — satisfies Definition (2.2) of a Markov chain
is not obvious, and a proof can be found in

To state the converse of the foregoing, if X,, is a Markov chain with transition probabilities
p,, and initial distribution p, then its finite dimensional distributions are given by (2.3). Proof of
this is also found in

Now that we have put the theory on a firm, if abstract, foundation, we can bring the
discussion down to earth by making the forgoing a little more concrete. First, we specialize our
study of Markov chains by assuming that our chain is temporally homogeneous, which means that
the transition probabilities do not depend on time; i.e. p,,(w,,, B) E p(w,., B). (This is the stochastic
analogue of a conservative system.) Next we assume that our state space S is finite, and suppose
for all states 1', j E S that p(i, j) Z 0, and E, p(z', j) = 1 for all i. In this case, equation (2.2) takes

a more intuitive form:
P(Xn+l =j I X7: = 7:) =P(iaj)

and our transition probabilities become
% 08.txt
If P is a matrix whose (i, j)th element is the transition probability p(i, 1') then P is a stochastic

matriar, that is, a matrix with elements pgj satisfying

Pij 20: zjpij =1: (i7j=1121"‘)d)

We also refer to P as the transition probability matrix.

Without loss of generality, we can further suppose our Markov chain is irreducible. This
means that, for any states 2', 1‘, starting in state i the chain will make a transition to state 1' at
some future time with positive probability. This state of affairs is often described by saying that
all states communicate. We lose no generality with this assumption because any reducible Markov
chain can be factored into irreducible classes of states which can each be studied separately.

The final two conditions we place on the Markov chains considered below will cost us
some generality. Nonetheless, there remain many examples of chains meeting these conditions,
thus making the present study worthwhile. Furthermore, it may be the case that, with a little
more work, we will be able to drop these conditions in future studies. The first condition is that
the chain is aperiodic. If we let I, = {n 2 1 : p"(:c,a:) > 0}, we call a Markov chain aperiodic if,
for any state 2:, the greatest common divisor of I, is 1. The second assumption is that our chain is

reversible. This characterization is understood in terms of the following definition.
Definition 2.1.3 A measure a is called reversible if it satisfies

;4(w)p(I. y) = u(y)p(y. w)

for all :c and y.

We call a Markov chain reversible if its stationary distribution (defined in Section 2.1.2) is reversible.
% 09.txt
\subsection{A Convergence Theorem}

In succeeding arguments, we use some results concerning the asymptotic behavior of

Markov diains. These results require a few more definitions.

Definition 2.1.4 A measure 1r is said to be a stationary measure if

Z«(z)p<z.y> = we) (2.4)

2
Equation 2.4 says P,,(X1 = y) = 1r(y), and by induction that P,,(X,, = y) = 7r(y) for all n 2 1. If 1r
is a probability measure, then we call 1r a stationary distribution. It represents an equilibrium for
the chain in the sense that, if X0 has distribution 1r, then so does X,. for all 11..

When the Markov chain is irreducible and aperiodic, the distribution of the state at time
n converges pointwise to 1r as n —> oo, regardless of the initial state. It is convenient to state
this convergence result in terms of the Markov chain's transition probability matrix P. Before
doing so, we note that irreducibility of a Markov chain is equivalent to irreducibility (in the usual
matrix theory sense) of its transition probability matrix. Furthermore, it turns out that a transition
probability matrix of an aperiodic Markov chain falls into that class of matrices often called acyclic,
but for simplicity we will call such stochastic matrices aperiodic. With this terminology, we can

state the convergence theorem in terms of the transition probability matrix P.

% Theorem 2.1.1
\begin{theorem}
\label{thm-2.1.1}
Suppose P is irreducible, aperiodic, and has stationary distribution 7r. Then as n —> oo,
p"(i,J') -> #0")
\end{theorem}

The symbol p"(i, j) is to be understood as the (ij) element of the nth power of P.
A Markov chain whose transition probability matrix satisfies the hypotheses of
Theorem~\ref{thm-2.1.1} is called ergodic. If we simulate an ergodic chain for
sufficiently many steps, having 
% 10.txt
begun in any initial state, the final state is a sample point from a distribution which is close to 1r.

To make this statement more precise requires that we define “close”.

Definition 2.1.5 Let 1r be a probability measure on S. Then the total variation distance at time

n with initial state I is given by
AA") = l|P"(z, A) - 7r(-4)||Tv = 1333; I P"(I.A) - 7r(A) I

In what follows, we will measure rate of convergence using the function 'r,(e), the first time after

which the total variation distance is always less than 5. That is,
-r,,(e) = min{m : A,(n) 3 E for all n. 2 m}

To consider the connection between convergence rates and eigenvalues, note that an ape-
riodic stochastic matrix P has real eigenvalues 1 = /\o > A1 2 A2 3  Z /\¢_1 2 -1, where
d = |S'] is the dimension of the state space. For an ergodic chain, the rate of convergence to the
stationary distribution 7r is bounded by a function of the subdominant eigenvalue. By subdominant
eigenvalue we mean that eigenvalue which is second-largest in absolute value, and we denote this
Am“ E max{).1,|A¢_1|}. The function bounding the rate of convergence is given by the following

theorem:

% Theorem 2.1.2
\begin{theorem}
\label{thm-2.1.2}
The quantity T1(€) satisfies
1. 13(5) 5 (1 — /\m,_x)“(ln1r(a:)‘1 +1ne‘1)
2. maxxes r,(e) 2 %/\,,,,x(1 — ).,,,,x)‘1ln(2e)’1
\end{theorem}

As this theorem shows, if we have an upper bound on the subdominant eigenvalue, then we have
an upper bound on the function 11(5). In what follows, we will derive an approximation to the
% 11.txt
subdominant eigenvalue and supply error bounds. Together, an approximation and error bounds
for Am“ provide enough information to make Theorem~\ref{thm-2.1.2} useful.

\section{Functions on the State Space}

Recall that X,.(w) = w,. 6 S denotes the state in which the Markov chain exists at time
11. Suppose that 11> = {¢1, . . . , ¢,,} is a collection of p observables, or functions defined on the state
space S. Furthermore, let these observables be real valued, so ¢.~ : S —> R. It is often useful, as we
will see below, to assume that none of the observables is a constant function. Suppose now that
the state space S is finite with at possible states. Then, since an observable is simply a map of the
state space, we can think of each ¢,- as a vector of d real numbers — the d values that it takes on at
the different states.

Now assume the Markov chain is irreducible, and let 1r denote its stationary distribution.
If we start the chain from its stationary distribution - i.e. suppose X0 has distribution 1r - then
X,. is a stationary process. Furthermore, for each i, ¢,(X,.) is a stationary stochastic process with

mean

E.¢.- = Z) vr(x)¢.(z)

165

and autocovariance function

Cir (¢i(Xn)v ¢i (Xn+a))

E7!’  ‘ E1r¢i) (¢i(Xn+a) " E1r¢i)l

Z Pn(Xn = z1X7I+S = y)  — E1r¢i)  _ E1r¢i)

z,yES
(2.5)

which, by the definition of conditional probability, we can write as,

2 P«(X.. = z)P.(X..+.. = y I X. = 1)(¢s($) — Em) (my) ~ Em)
1.1165
% 12.txt
Z 7r($)P1y (MI) - Em.-) (¢:(1I) - E1r¢i)

:c,y€S

Here pgy denotes the element in row :1: and column y of P’, the sth power of the transition probability

matrix. Similarly, we define the cross-covariance between the function ¢,- at time n and q5,- at time

n+sas

E1!’  _ E7r¢:') (¢i(Xn+a) " E1r¢i)l (2-6)

C1r(¢i(Xn)a (#1 (Xn+a))

Z) vr(x)p;,, (¢.-(w) — Em.-) (my) — Em)

z,yES

Now let (<I>) denote the matrix of mean vectors whose jth column is E,,¢,-1, where 1 = (1, . . . , 1)‘,
and let H = diag(1r(w1), . . . , 1r(wd)), be the d X d diagonal matrix with stationary probabilities 1r(w)
on the main diagonal, and zeros elsewhere. Finally, denoting by C(s) the p X p covariance matrix

whose (i,j)th element is C, (qfi,-(X,.),¢_,< (X,.+,)), we have

0(0) = E(‘1’(Xn) - (‘I>))(‘I’(Xn) - (‘PW
= (‘P-(‘1’))‘1'1(<1’-(‘I>))
0(8) = E(<I>(Xn) — (‘I>))(‘1’(Xn+a) - (‘1’))’

(‘P - (‘1’))'1'IP‘(<1> — (<1>))

Below, we will also find it useful to have at our disposal a new matrix that is similar to the
transition probability matrix. We have in mind the matrix M E Ill/2PI'I‘1/2. As is easily verified,
this allows us to write the covariance matrix as

can = <<I>—<<I>>)°n%*M*n%<<I>—<<I>>>

\II‘M’\II (2.7)

where we have defined ‘II E 111/ 2(<I> —  Recall that our main motivation is that, for out-of-core
problems, traditional eigenvalue algorithms are inadequate. With this fact, and the form (2.7) in
% 13.txt
mind, we will consider using the covariance of observables on the state space to implement the
(otherwise useless) Rayleigh-Ritz procedure, which we describe below. This procedure requires
that M be symmetric. As the next fact demonstrates, this required symmetry is the reason for our

interest in the reversibility property of the Markov chain.

Fact 2.2.1 The matrix M is symmetric if and only if the Markov chain is reversible (i.e. iﬂ the

process satisfies the detailed balance condition).

Proof:

M is symmetric

4:, (1-I1/2P1-I-1/2): = H1/2Pn—1/2
<.—> r1-%*P°n%* = H1/2911-1/2
¢> 1>'r1= = up
~:=> (IIP)‘ = l'IP

Element-wise, the final equality implies that 7|‘,'p,'j = 1rjpj.-. According to Definition 2.1.3, this states

that p is a reversible measure.

% 14.txt

@@

% 22.txt
\chapter{Lanczos Procedures}

The following will review the usual procedure for generating an orthonormal basis for
lC(M, «/11, p), and conclude by showing that, for our problem, we can find the matrix Q‘MQ without
actually carrying out this procedure. Note that this conclusion is essential since the procedure

requires matrix vector multiplication - an operation we have assumed impossible for large enough :1.

4.1 The General Lanczos Algorithm
Let 21;“) E M"¢ and consider the d x :1 matrix
«I 2 [«/»,M«/»,  ,M"“¢] 2 [«b‘°>,«/»“’,...,¢<‘-‘>1

Notice that M\II = [1/:(1),... ,1/;(d'1),Md¢r] and, assuming for the moment that \I1 is non-singular,

we can compute the vector h = —\II‘1M"1/J. Thus,

M\II = \II[e2,. . . ,e,1, —h] E IIIH

% 23.txt
23

Equivalently,
0 0  0 -—h1
1 0  0 —h2
HE   3 ; =\II“M\II (4.1)
0 "hl1—1
1 —h,;

H is a companion matrix which means that its characteristic polynomial is p(:r) = 2:‘ + 23:1 h,-:r"‘1.
Since H is similar to M, ﬁnding the eigenvalues of M is equivalent to ﬁnding the roots of p(a:).
However, this is of little practical use since ﬁnding h, constructing 12(1), and ﬁnding its roots is
probably a harder problem than the one we started with. Instead, the value of decomposition (4.1)
derives from its upper Hessenberg form. We exploit this property below.

Let ‘II = QR be the QR decomposition of \II. Since ‘I! is assumed nonsingular,
~II;‘M~r.: = (R"Q‘)M(QR) = H
Therefore,
Q‘MQ = RHR-1 2 T

Since R and R“1 are both upper triangular and H is upper Hessenberg, T = R.HR'1 is also upper
Hessenberg. Furthermore, since M is symmetric, it is clear that T‘ = Q‘MQ = T. That is, T is

an upper Hessenberg matrix that is symmetric. Therefore, T is tridiagonal and we can write it as

follows:

T =    (4.2)

ﬁd—1

0  174-1 ad

% 24.txt
24

Equating columns j on both sides of MQ = QT yields
M113‘ = B,-—1qj—1 + 0:91 + ﬁjqj+1 (43)

Since the columns of Q are orthonormal, multiplying both sides of (4.3) by q_,- and q,-+1 yields
oz, = q§Mq,- and ﬂj = q;-+1Mq,-.
The foregoing justifies what is called the Lanczos algorithm, performed as follows:

The Lanczos algorithm for (partial) reduction to symmetric tridiaganal form

91 = ¢/||¢||2, [30 = 0, 40 = 0

for j=1top

z=Mq_,-

5:‘ = ||Z||2
if 5, = 0, quit
¢Ij+1 = 2/51

end for

The qj computed by the Lanczos algorithm are often called the Lanczos vectors. If the
loop in the algorithm is terminated because ﬁg = 0, this indicates that an exact invariant subspace
has been computed, and is given by ran{q1 - - - qk}. Otherwise, we usually halt the algorithm after
p steps, in which case the algorithm converg to approximations of at most 12 eigenvalues.‘

Note that equation (4.3) results from the equation MQ = QT. The latter equation holds
since, starting with d vectors in our subspace, Q is a d Xd matrix whose columns form an orthonormal

basis for all of R‘, which is clearly an invariant “sub” space. The Lanczos algorithm above, however,

‘A detailed discussion of such convergence (and misconvergence) is given in

% 25.txt
25

proceeds for only p steps, producing a d x p matrix Q, whose columns form an orthonormal basis
for the approximate invariant subspace IC(M, ¢, p), and a p X p matrix T, = Q'MQ. In that case,
IIMQ = QTp|[g =  is non-zero and gives the error bound described in Theorem 3.2.1. Now,

writing the full d x :1 matrix T of (4.2) as

01 51
/71

(4.4)

allows us to describe the error bound ||E||-2 in terms of the submatrix Tm, and, hence, in terms of

ﬁzz-
Theorem 4.1.1

HT, and T,“ are the matrices appearing in (4.4), and if the p columns of Q are computed by the

Lanczos algorithm, then there exist pl, . . . ,;t,, E /\(T,) and A1, . . . , A, E A(M) such that

ill» - /\I=i S iiTP'4”2 = 512

fork=1,...,p.

For a proof, see page 365 of [1].

% 26.txt
26

4.2 A Lanczos Procedure for Markov Chains

Notice that the Lanczos algorithm requires the computation of Mqj. Throughout this
paper we have assumed that M 6 Rd“, and that d is so large as to make the matrix vector
multiplication Mqj impossible. Now, the Lanczos algorithm constructs an orthonormal basis for
the Krylov subspace. For our purposes, this construction is not essential. What is essential is that
we ﬁnd a way to generate the Lanczos coejﬁcients oz,-, ﬁj, (j = 1, . . . , p), and whence the matrix
T, without performing the operation Mqj required by the Lanczos algorithm. We now address this
problem.

Begin with the centered and weighted observable 1/1 = 111/ 2 (qS—E,,¢), and let q1 = 11;/||1/JH2.
To simplify notation, let ¢(n) E ¢(X,.). The form of the first Lanczos coeﬂicient a1 is straight

forward:

01 = qiMq1

I/1‘M1/I
¢’¢

and now the special properties of the Markov chain setting play an important role. For, regard the

following

¢‘M«/2 = (¢ — E.¢)‘II%‘MII%(¢ - Em)

Cw (d>(n), ¢(" + 1)) (4-5)

and

W1!’ = (<15 - E«¢)'H(¢ - E«¢)

Varn (¢(")) (4-6)

Recall that the deﬁnition of M implies H%‘MI'I% = HP, and this justiﬁes equation (4.5). Putting

% 27.txt
27

these equations together, we have a. nice form for :11:

_ c, (run), an + 1))
"‘ ' Ta:,(¢(n))

The next Lanczos coeﬂicient, B1, is only slightly more complicated.

[31 = ||M91-0t1tI1||2

[(Mq1 — 0l1¢11)i(M¢]1 - 0101)] 1/2

[t1iM2¢I1 — ail 1/2

(4.7)

Again recalling the deﬁnition M E 111/ 2P1'I‘1/2, a. simple manipulation veriﬁes

1'li‘M’l'[% = HP’, whereby it follows that

(«s — E,,¢)‘HP”(¢ — Em)

"‘M2"‘ (¢ — E,¢)*n<¢ — Em)

C: (¢("), ¢(n + 2))
V3-Ur

Inserting the last expression into equation (4.7) gives

_ c,(¢(n).¢<n+2)) c,(¢<n>,¢(n+1)> ’ "’
"“ jvar,<¢<n)> "(jvar,<¢<n)) ‘)

Notice that we can estimate these expressions of oz; and ﬂl by running simulations of the Markov
chain. To do so, we run the Markov chain starting from a random initial state, compute the value of
the observable at each step of the chain and, from the data, compute estimates of the covariances
and variance of the observable.

Before generalizing this for the jth Lanczos coefficient, we consider one more coeﬂicient
in detail. Recall that the second basis element q; in the Lanczos algorithm is given by 1;; =

(Mql — oz1q1)/ﬁl. From this we can derive an expression for (12 which, though not as neat as those

% 28.txt
28

for oz; and [31, is nonetheless tractable.

I12 ‘IiMq2

1
[7(M91 — 0t1¢I1)tM(M¢I1 - 11191)
1

1
= f,§(qiM3q1 - 2a1qlM2<11 + a’9iM¢11)
1
1
= —2(<1iM3q1 - 2015i ‘ all (4-8)

1.
The last equality follows from equation (4.7), which can be written as q{M2q1 = off’ + (if. The
preceding paragraphs deﬁne all but one of the quantities in the expression (4.8). The one remaining
is q§M3q1, which can also be expressed in terms of covariances of <15, and the analysis is almost

identical to that for q{M2q1:

(¢ — E,¢)'IIP3(¢ — Em)

"lM3"‘ (¢ — E,¢s)‘H(¢ - Em»)

Cu (¢(")a ¢(n + 3))
Varn(¢(n))

To generalize the foregoing for higher order Lanczos coeﬂicients requires a recurrence
relation between (aj+1, ﬂj) and (aj, ﬂ,~_1). As we will see, such a relation involves higher order
Lanczos vectors q,~ — vectors we do not wish to compute. Recall that the only “vector” we can
handle is our initial observable. Therefore, we must also establish a recurrence among the q,-’s, thus
making them available inductively from our observable vector.

To begin with the ﬂfs, recall that ﬁj E ||Mqj — ajq_.,- — ,H_,~_1q,-_;|[2. Whence,

Bf (M411 — °‘j‘Ij - ﬁj-1qj—1)’(MlIj - Otjqj - 51-19141)

q;-M291‘ - 0? - B,z_1 (4-9)

Some simple algebra and the deﬁnitions aj E q§Mqj and ﬁJ~_1 E q;-Mq,--1 prove equality (4.9). The
% 29.txt
recurrence for 0114.1 is equally straight forward, though the ﬁnal form is not as neat:

2
0tj+1 ¢Ij+1M'1j+1

1
§(M¢15 - 11111;‘ - ﬂj—1<Ij—1)'M(Mqj - Otjqj - /3j—1qj—1)
J

1
§(q;-M-“*4; — 2a,~q;M*q,- — 2x3;+’_1q;--1M*q,- + «I? + 2a,-5,51 + am (4.10)
J
From (4.9), q;~M2q,~ = oz? + ﬂf + ﬂ}_1 and, upon plugging this into (4.10), we arrive at
1
a,~+1 = E(q§»M3q_,,- — (1? — 2ozJ~ﬁJ2 — 2ﬂj_1q;»_1M2qj + ﬁ2_1) (4.11)
J

The expressions above involve the higher order Lanczos vectors qj. In the present context,
these vectors will not be readily available. However, another recursion — this time relating forms
involving q,-+1 to forms involving qj and qj__1 — will make these expressions accessible by induction
on our initial observable. The general forms of interest are q;+1M"qj+1 and qj-+1M"qJ~. Again, we

begin with the simpler of the two:

¢I;'+1Mk9j = ij(Mqj - Otjqj - ﬂj—1t1j—1)'M"¢1j
= i(<1§M"“¢Ij - C‘j‘I_;'Mk‘Ij - ﬂj—1q§M"qj-1)
The second is,
‘I_;'+1Mkqj+1 = ﬂi?(M<Ij - Otjqj - .3j—1<Ij—1)‘Mk(Mqj — Orjqj - ﬂj—1¢1j—1)

1
ﬁ(‘1§M'°"<1:‘ + °‘:2"I§M'°<J:' + ﬁ?—z<Ii—1M"<I:‘—1
J
—2a,-q;-M*+‘q,~ + 2am,--1q;-M*q,--1 — zza,--1q;-M*+‘q]--1) (4.12)

>>>>

@@

% 22.txt
